Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions

### 2021-08-26 analysis/paper/shceof.Rmd
Muchos cambios

coord_polar <- function(ymax = -20) {

@@

  cbind(x, y) |> 
    list() |> 
    sf::st_polygon() |> 
    sf::st_sfc(crs = "+proj=latlong") -> white

  list(
    geom_sf(data = white, inherit.aes = FALSE, 
            fill = "white", 
            colour = "white", size = 2),
    coord_sf(ylim = c(-90, ymax), 
             lims_method = "box",
             crs = "+proj=laea +lat_0=-90",
             default_crs = "+proj=longlat",
             label_axes =  "----")
  )
}

@@

```{r read-data}
era5 <- ReadNetCDF(ERA5(), vars = c(hgt = "z",
                                    t = "t"),
                   subset = list(level = list(850, 700, 200, 50),
                                 latitude = -90:10)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w), 
        t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))]
```

```{r compute-psi}
# Scaling magical numbers. (come from psi-test.Rmd)
# Not really important, though.
scaling <- c(add = -2.364646e+06, mult = 3.954828e+13)

psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(vo = mean(vo*w)), by = .(lon, lat, lev, time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  setnames("value", "psi") %>% 
  .[, psi := psi*scaling[2] + scaling[1]] %>%
  .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
```

```{r o3}
o3 <- O3() %>% 
  ReadNetCDF(vars = c(toc = "tco3")) %>% 
  normalise_coords() %>% 
  .[, lon := ConvertLongitude(lon)] %>% 
  .[lat <= -20] %>% 
  .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
  .[, w := season_weights(time)] %>% 
  .[, .(toc = mean(toc*w)), by = .(lon, lat, time = seasonally(time))]

@@

```{r psa}
psa <- ReadNetCDF(ERA5(), vars = c(hgt = "z"),
                  subset = list(level = list(500),
                                latitude = -90:0)) %>% 
  normalise_coords() %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))] %>% 
  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), 
    by = .(lon, lat, season(time))] %>%
  EOF(value ~ time | lon + lat, n = 1:3, data = .)

@@

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two PSA [@mo2001].
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

@@

```{r compute-ceof}
invisible(
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = -90:10,
                           level = list(200, 50))) %>%
    normalise_coords() %>%
    .[season(time) == "SON"] %>%
    .[, w := season_weights(time)] %>% 
    .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
    .[, season := season(time)] %>%
    {
      ceof <<- .[, .(eof = list(compute_ceof(.SD))), by = .(season)]
      ceof_splitted <<-  .[, levs := lev] %>%
        .[, .(eof = list(compute_ceof(.SD, n = 1:3))), by = .(season, levs)]

@@

table2 <- table %>%
  widyr::pairwise_cor(item, time, mag) |> 
  setDT() %>%
  .[, p.value := correlation_statistics(correlation, n)$p.value] %>% 
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)]

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]

@@

```{r compute-regression}
regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
    by = .(season, cEOF)] %>%
  # dcast(time ~ cEOF, value.var = c("Real", "Imaginary")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, Imaginary, Real, se = TRUE),
    by = .(lev, lon, lat, variable, cEOF)] %>%
  # .[, c("term", "cEOF") := tstrsplit(term, "_")] %>%
  rm_intercept() %>%
  .[, term := factor_ReIm(term)]
```

@@

mgp <- latex2exp::TeX(r'(Geopotential height ($m^2s^{-1}$))')
```

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

@@

sam_r2 <- sams %>%
  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]

@@

```{r t}
t <- ERA5() %>% 
  ReadNetCDF(vars = c("t", hgt = "z", "o3"), 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(t   = mean(t*w),
        o3  = mean(o3*w),
        hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
```

@@

```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=2.5, fig.width=width_column}
t %>% 
  .[lat %between% o3wave_lats] %>% 
  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
        o3 = weighted.mean(o3, cos(lat*pi/180)),
        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  melt(id.vars = c("lev", "time", "lon")) %>% 
  .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(cEOF)] %>% 
  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  ggplot(aes(lon, lev)) +
  geom_contour_fill(data = ~.x[variable == "t"],
                    aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
  geom_contour_tanaka(data = ~.x[variable == "t"],
                      aes(z = estimate),
                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
  geom_contour2(data = ~.x[variable == "o3"], 
                aes(z = estimate, linetype = factor(-sign(..level..))), 
                breaks = AnchorBreaks(0, exclude = 0)) +
  geom_text_contour(data = ~.x[variable == "o3"], 
                    aes(z = estimate*1e9, label = signif(..level.., 2)),
                    size = 2,
                    skip = 1, stroke = 0.15, stroke.colour = "white", 
                    label.placement = lable_placement_min(),
                    breaks = AnchorBreaks(0, exclude = 0)) +

@@

```{r regr_psi}
regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
    by = .(season, cEOF)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
  rm_intercept() %>%
  .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, cEOF, season, term)] %>%
  .[, term := factor_ReIm(term)]
```

@@

```{r plot_sst_psi}
plot_sst_psi <- function(which_cEOF, which_lev) {
  sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
      by = .(cEOF, term)] %>%
    .[, var := "SST"]

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) 
```

@@

```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200) 
```

@@

```{r cmap}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, w := season_weights(time)] %>% 
  .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
```

```{r regr_pp}
regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
    by = .(season, cEOF)] %>%
  .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
  .[season == "SON"] %>%
  .[, FitLm(value, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, variable, season)] %>%
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)]
```

@@

@@

```{r plot_pp}
plot_pp <- function(data, which_cEOF, which_term = c("Real", "Imaginary"), which_continent = NULL) {
  filter_continent <- function(data) {
    if (!is.null(which_continent)) {
      data[continent == which_continent]
    } else {
      data
    }

  }
  data %>%
    .[cEOF %in% which_cEOF] %>%
    .[term %in% which_term] %>%
    .[, lev := "sfc"] %>%
    .[lat >= -60] %>%
    add_continent() %>%
    filter_continent() %>%
    # .[abs(estimate) <= 2] %>%
    .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
      by = .(lev, cEOF, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    # geom_raster(aes(fill = estimate)) +
    geom_contour_fill(aes(z = estimate, fill = ..level..), 
                      breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +

    # scale_fill_divergent() +
    scale_fill_divergent_discretised("Precipitation (mm/day)", 
                                     # limits = c(-.8, .8),
                                     # oob = scales::squish,
                                     low = "#8E521C",
                                     high = "#00665E",
                                     guide = guide_colorsteps_bottom(15)) +
    # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
    geom_contour_pval(aes(z = p.value)) +
    geom_qmap(keep = 1) +

    scale_x_continuous("", expand = c(0, 0), breaks = seq(0, 360, by = 15), 
                       labels = LonLabel) +
    scale_y_latitude(ticks = 15) +
    ggh4x::facet_nested(term ~ cEOF) +
    axis_labs_smol +
    tag_facets("rc") +
    coord_sf(ylim = continents[[which_continent]]$lat,
             xlim = continents[[which_continent]]$lon,
             crs = "+proj=longlat +pm=180 +over",
             default_crs = "+proj=longlat +over") +
    theme(panel.spacing = grid::unit(1, "lines"))

>>>>

coord_polar <- function(ymax = -20, ...) {

@@

  cbind(x, y) %>% 
    list() %>% 
    sf::st_polygon() %>% 
    sf::st_sfc(crs = "+proj=latlong") -> white

  list(
    geom_sf(data = white, inherit.aes = FALSE, 
            fill = "white", 
            colour = "white", size = 2),
    coord_sf(ylim = c(-90, ymax), 
             lims_method = "box",
             crs = "+proj=laea +lat_0=-90",
             default_crs = "+proj=longlat",
             label_axes =  "----", ...)
  )
}

@@

```{r read-data}
era5 <- ERA5_geopotential() %>% 
  ReadNetCDF(vars = c(hgt = "z"),
             subset = list(level = list(200, 50),
                           latitude = -90:10)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
```

```{r compute-psi}
# Scaling magical numbers. (come from psi-test.Rmd)
# Not really important, though.
scaling <- c(add = -2.364646e+06, mult = 3.954828e+13)
psi <- ERA5_vorticity() %>% 
 ReadNetCDF(vars = "vo") %>%
  normalise_coords() %>%
  .[, lev := 200] %>% 
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(vo = mean(vo*w)), by = .(lon, lat, lev, time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  setnames("value", "psi") %>% 
  .[, psi := psi*scaling[2] + scaling[1]] %>%
  .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
```

```{r o3}
o3 <- ERA5_TOC() %>% 
  ReadNetCDF(vars = c(toc = "tco3")) %>% 
  normalise_coords() %>% 
  .[lat <= -20] %>% 
  .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
  .[, w := season_weights(time)] %>% 
  .[, .(toc = mean(toc*w)), by = .(lon, lat, time = seasonally(time))]

@@

```{r psa}
psa <- ERA5_geopotential() %>% 
  ReadNetCDF(vars = c(hgt = "z"),
             subset = list(level = list(500),
                           latitude = -90:0)) %>% 
  normalise_coords() %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))] %>% 
  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), 
    by = .(lon, lat, season(time))] %>%
  EOF(value ~ time | lon + lat, n = 1:3, data = .)

@@

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two PSA [@mo2001].
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

@@

```{r compute-ceof}
invisible(
  ERA5_geopotential() %>% 
    ReadNetCDF(vars = c(hgt = "z"),
               subset = list(latitude = -90:10,
                             level = list(200, 50))) %>%
    normalise_coords() %>%
    .[season(time) == "SON"] %>%
    .[, w := season_weights(time)] %>% 
    .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
    .[, season := season(time)] %>%
    {
      ceof <<- .[, .(eof = list(compute_ceof(.SD))), by = .(season)]
      ceof_splitted <<-  .[, levs := lev] %>%
        .[, .(eof = list(compute_ceof(.SD, n = 1:3))), by = .(season, levs)]

@@

table2 <- table %>%
  widyr::pairwise_cor(item, time, mag) %>% 
  setDT() %>%
  .[, p.value := correlation_statistics(correlation, n)$p.value] %>% 
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)]

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5_geopotential(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]

@@

```{r compute-regression}

era5 <- ERA5_temperature() %>% 
  ReadNetCDF(vars = c("t"), 
             subset = list(level = list(50, 200))) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))] %>% 
  .[era5, on = .NATURAL]

regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
    by = .(season, cEOF)] %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[, FitLm(value, Imaginary, Real, se = TRUE),
    by = .(lev, lon, lat, variable, cEOF)] %>%
  rm_intercept() %>%
  .[, term := factor_ReIm(term)]
```

@@

mgp <- latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
```

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

@@

sam_r2 <- sams %>%
  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>% 
  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]

@@

```{r t}
t <- ERA5_temperature() %>% 
  ReadNetCDF(vars = c("t"), 
             subset = list(latitude = o3wave_lats)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>% 
  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lon, lev, time)]

o3_temp <- ERA5_ozone() %>% 
  ReadNetCDF(vars = c("o3"), 
             subset = list(latitude = o3wave_lats)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>% 
  .[, .(o3 = weighted.mean(o3, cos(lat*pi/180))), by = .(lon, lev, time)]

t <- t[o3_temp, on = .NATURAL]
rm(o3_temp)

t <- t %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(t   = mean(t*w),
        o3  = mean(o3*w)), by = .(lev, lon, time = seasonally(time))]
```

@@

```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=2.5, fig.width=width_column}
t %>% 
  melt(id.vars = c("lev", "time", "lon")) %>% 
  .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(cEOF)] %>% 
  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  ggplot(aes(lon, lev)) +
  geom_contour_fill(data = ~.x[variable == "t"],
                    aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
  geom_contour_tanaka(data = ~.x[variable == "t"],
                      aes(z = estimate),
                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
  geom_contour2(data = ~.x[variable == "o3"], 
                aes(z = estimate, linetype = factor(-sign(..level..))), 
                breaks = AnchorBreaks(0, exclude = 0)) +
  geom_text_contour(data = ~.x[variable == "o3"], 
                    aes(z = estimate*1e9, label = signif(..level.., 2)),
                    size = 2,
                    skip = 1, stroke = 0.15, stroke.colour = "white", 
                    label.placement = lable_placement_min(),
                    breaks = AnchorBreaks(0, exclude = 0)) +

@@

```{r regr_psi}
regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
    by = .(season, cEOF)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[lev == 200] %>% 
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
  rm_intercept() %>%
  .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, cEOF, season, term)] %>%
  .[, term := factor_ReIm(term)]
```

@@

```{r plot_sst_psi}
plot_sst_psi <- function(regr_sst, regr_psi, which_cEOF, which_lev) {
  sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
      by = .(cEOF, term)] %>%
    .[, var := "SST"]

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
             which_cEOF = "cEOF2", which_lev = 200) 
```

@@

```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
             which_cEOF = "cEOF1", which_lev = 200) 
```

@@

```{r cmap}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -60:10)) %>%
  normalise_coords() %>%
  .[, w := season_weights(time)] %>% 
  .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
```

@@

crop_continent <- function(continent) {
  bbox <- c(xmin = min(continents[[continent]]$lon), 
            xmax = max(continents[[continent]]$lon),
            ymin = min(continents[[continent]]$lat),
            ymax = max(continents[[continent]]$lat))

  function(data) {
    s2 <- sf::sf_use_s2(FALSE)
    on.exit(sf::sf_use_s2(s2))
    data <- sf::st_crop(data, bbox)
    data$continent <- continent
    data
  }
}
```

```{r}
rotate_reim <- function(real, imaginary, angle = 0) {
  z <- complex(real = real, imaginary = imaginary)
  z <- rotate(z, angle)
  list(real = Re(z),
       imaginary = Im(z))
}

sst_for_regression <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit()

angles_regr <- -c(0, 45, 90, 45+90)*pi/180

sst_regr <- lapply(angles_regr, function(a) {
  sst_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a]
}) %>% 
  rbindlist()

psi_for_regression <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[lev == 200]

psi_regr <- lapply(angles_regr, function(a) {
  psi_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
    .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), 
      by = .(lev, cEOF, season, term)] %>%
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a]
}) %>% 
  rbindlist()

pp_for_regression <- ceof[, eof[[1]]$left, by = season] %>%
  .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[cmap, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit() %>% 
  .[, pp := pp/sd(pp), by = .(lon, lat)]

pp_regr <- lapply(angles_regr, function(a) {
  pp_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(pp, Imaginary, Real, se = TRUE),
      by = .(lon, lat, cEOF, season)] %>%
    .[term == "Real"] %>% 
    rm_intercept() %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a*180/pi]

}) %>%
  rbindlist()

# ceof[, eof[[1]]$left, by = season] %>%
#   .[cEOF == "cEOF2"] %>% 
#   copy() %>% 
#   sep_ReIm(hgt, "wide") %>%
#   .[, hgt := NULL] %>%
#   .[, on = "time", allow.cartesian = TRUE] %>%
#   .[season(time) == "SON"] %>% 
#   na.omit() %>% 
#   .[, pp := pp/sd(pp), by = .(lon, lat)]

@@

```{r plot_pp}
spoke <- function(data, coords) {
  radius <-  grid::unit(.15, "snpc")

  outer_circle <-  grid::circleGrob(x = 0, 
                             y = 0,
                             r = radius*1.2)

  xcentre <- grid::unit(0, "npc") + grid:::grobHeight(outer_circle)/2
  ycentre <-  grid::unit(0, "npc") + grid:::grobWidth(outer_circle)/2

  circle <- grid::circleGrob(x = xcentre, 
                             y = ycentre,
                             r = radius, 
                             gp = grid::gpar(fill = "white", 
                                             alpha = 0.7))
  # browser()
  dx <- cos(coords$angle[1]*pi/180)*grid::grobHeight(circle)/2
  dy <- sin(coords$angle[1]*pi/180)*grid::grobWidth(circle)/2

  line <- grid::segmentsGrob(x0 = xcentre, y0 = ycentre, 
                             x1 = xcentre + dx,
                             y1 = ycentre + dy, 
                             gp = grid::gpar(fill = "black"),
                             arrow = grid::arrow(angle = 13, 
                                                 length =  grid::unit(.15/2, "snpc"),
                                                 type = "closed")
  )

  grid::gTree(children = grid::gList(circle, line))

}

plot_pp <- function(data, which_cEOF) {
  lats <- c(-55, -10)
  breaks <- AnchorBreaks(exclude = 0)(c(-.9, .9))

  data %>%
    # .[angle == unique(angle)[2]] %>% 
    .[lat %between% lats] %>% 
    # .[cEOF %in% which_cEOF] %>%
    # .[term %in% which_term] %>%
    .[, lev := "sfc"] %>%

    .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
      by = .(lev, cEOF, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +

    geom_contour_fill(aes(z = estimate, fill = ..level..),
                      breaks = breaks) +
    geom_contour_tanaka(aes(z = estimate), breaks = breaks) +

    scale_fill_divergent_discretised("Precipitation (correlation)", 
                                     # limits = c(-1, 1),
                                     # oob = scales::squish,
                                     low = "#8E521C",
                                     high = "#00665E",
                                     guide = guide_colorsteps_bottom(15)) +

    geom_contour_pval(aes(z = p.value)) +
    geom_qmap() +

    grid_panel(spoke, aes(angle = -angle)) +

    scale_x_continuous("", expand = c(0, 0), breaks = seq(0, 360, by = 15), 
                       labels = LonLabel) +
    scale_y_latitude(ticks = 15) +
    facet_grid(angle ~ cEOF) +
    axis_labs_smol +
    tag_facets("rc") +
    coord_sf(ylim = lats, 
             crs = "+proj=longlat +pm=180 +over",
             default_crs = "+proj=longlat +over") +
    theme(panel.spacing = grid::unit(1, "lines"))
